---
title: "Data_Manipulation_Workingon092925"
output: github_document
date: "2025-09-30"
---


```{r}
library(tidyverse)
```

## Load the FAS litters data

```{r}
litters_df = read.csv("./data/FAS_litters.csv")
litters_df = janitor::clean_names(litters_df)
```


## `select`

Choose some columns and not others 

```{r}
select(litters_df, group, gd0_weight)
```

## Specifying the range of columns that I want to take a look at, we use colon :

```{r}
select(litters_df, group, gd0_weight:gd_of_birth)
```

## specifying what we want to get rid of, use minus sign

```{r}
select(litters_df, -litter_number )
```


Renaming columns

```{r}
select(litters_df, GROUP = group, LITTER_NUMBER = litter_number)
```

Another way of renaming - we can keep all the columns and then only change the name of selected ones by using rename(litters_df)

```{r}
rename(litters_df, GROUP = group, LITTER_NUMBER = litter_number)
```


HELPER FUNCTIONS - used to help accomplish but never use them on their own - select helpers 

Select helpers, columns that i know i want to keep: if i know the key phrases that exist in the beginning, end or middle of the variables then i can find them and select them

```{r}
select(litters_df, starts_with("gd"))
```

This here allows me to get litter_number at the beginning as the first variable and then everything else comes after - this is to be used when i know what column i want first

```{r}
select(litters_df, litter_number, everything())
```











